Estimating survival in left filtered data

### 2021-12-11 paper.tex
Interim results; running on GM data now

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Estimating survival in left filtered data: simulation and application},
  pdfauthor={Kevin Chen},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=2.54cm]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\PassOptionsToPackage{no-math}{fontspec}

@@

\title{Estimating survival in left filtered data: simulation and
application}
\author{Kevin Chen}
\date{Stat 256: Causal Inference (Fall 2021)}

@@

A central concern in the analysis of occupational cohorts is the
potential for the healthy worker survivor effect (HWSE), the phenomenon
by which healthy individuals remain at work, while less healthy
individuals leave work -- possibly in response to exposure-related
health decline. In the presence of the HWSE, those with the highest
cumulative occupational exposures are also those who are less at risk of
disease. Thus, standard measures of association would show an inverse
relationship between occupational exposure and poor health outcomes
(Arrighi and Hertz-Picciotto 1994). The HWSE is an example of
time-varying confounding affected by past exposure. Previous studies
have attempted to assess the presence of the HWSE in observed data by
assessing so-called path-specific associations using Cox proportional
hazards modeling (Naimi et al. 2013; Garcia et al. 2017). However, these
measures of associations are themselves subject to the confounding they
seek to quantify.

If sequential ignorability of exposure status at each point in follow-up
and positivity can be attained conditional on covariates, then causal
methods can be applied to account for the HWSE. Past studies have
applied causal methods capable of accounting for time-varying
confounding affected by past exposure to the study of MWF exposures and
cancer mortality outcomes in the UAW-GM Cohort Study (Garcia et al.
2018; Izano et al. 2019), but the study of cancer incidence outcomes is
further problematized by incomplete observation of cancer incidence
outcomes over the study period. We wish to make inferences about the
carcinogenicity of MWF exposure over an individual's lifetime starting
three years after they enter the workforce. The UAW-GM cohort includes
those hired roughly between 1938 and 1982. However, cancer incidence
reporting at the Michigan Cancer Registry did not begin until 1985. In
particular, we have the presence of \emph{left filtering} in the UAW-GM
Cohort Study when cancer incidence is the outcome of interest: before
1985, both cancer incidence status and time of cancer incidence are
unknown. Observation of the complete cancer incidence outcome vector
over the study period is conditional on an individual surviving to 1985
cancer-free.

In the presence of the HWSE, left filtering implies outcome
misclassification that is informative of true cancer status. As part of
her dissertation research, Izano (2017) conducted a quantitative bias
analysis for the estimation of survival curves in the presence of left
filtering and the HWSE. She simulated data compatible with the HWSE and
estimated cancer-free survival curves using an inverse probability of
treatment and censoring weighted Kaplan-Meier (WKM) estimator and the
WKM with an Aalen filter for left-filtering (AWKM) (Andersen et al.
1993; Xie and Liu 2005). Data were simulated under five different
scenarios. The causal estimands of interest were the survival curves
under interventions: (1) always exposed at work with no censoring due to
death and (2) never exposed at work with no censoring due to death. For
each intervention, three survival curves were produced: (1) the true
survival curve, (2) the WKM survival curve, and (3) AWKM survival curve.
The bias in mean survival difference calculated by AWKM was negligible
when the true 20-year risk of cancer was under 0.25.

Here, we replicate the simulation and bias analysis presented in Chapter
3 of Izano (2017), embed the problem in the non-parametric structural
causal approach of Pearl (1995), comment on the results, and apply the
WKM and AWKM estimators to the UAW-GM Cohort.

@@

The UAW-GM Cohort data includes person-year level exposure, outcome, and
covariate data starting three years after hire. To emulate the shape of
the data for this longitudinal cohort, we consider 20 years of data over
time indexed by years since hire. Notation representing the variables of
interest are presented in Table \ref{tab:variables}.

@@

Assume we have \(n = 50\,000\) iid units in \(X\) with
\[X_i(t) = \left(R_i = 0, W_i, S_i, \bar H_i(t), \bar N_i(t), \bar A_i(t), \bar a_i(t), \bar Y^*_i(t) = \bar Y_i(t) \right).\]
The bar notation to indicates variable history where
\(\bar X_i(t) = \left(X_i(k)\right)_{k = 1}^t\). In the case of
exposure, \(\bar A_i(t)\) is the indicator of whether unit \(i\) was
ever exposed through time \(t\). Note that true cancer status \(Y^*(t)\)
not observed until \(t \ge R\), after the start of the registry. Call
\(X\) the full data, where we have \(R = 0\) for all. In the observed
data \(X^\text{obs}\), we cannot assume \(R = 0\) for all, and
susceptibility \(S\) is not known:
\[X^\text{obs}_i(t) = \left(R_i, W_i, \bar N_i(t), \bar A_i(t), \bar a_i(t), \bar Y_i(t) \right).\]
Under the causal model, we assume the following non-parametric
structural equations: \[\begin{aligned}
R     & = f_R \left( U_R \right) \\
W     & = f_W \left( U_W \right) \\
S     & = f_S \left( U_S \right) \\
H(t)  & = f_{H(t)} \left( H(t - 1), U_{H(t)} \right) \\
N(t)  & = f_{N(t)} \left( W, N(t - 1), H(t), A(t - 1), U_{N(t)} \right) \\
A(t)  & = f_{A(t)} \left( W, \bar A(t - 1), N(t), U_{A(t)} \right) \\
D(t)  & = f_{D(t)} \left( W, \bar A(t - 1), D(t - 1), Y^*(t - 1), U_{D(t)} \right) \\
Y^*(t)& = f_{Y^*(t)} \left(W, S, H(t), \bar A (t), D(t), Y^*(t - 1), U_{Y^*(t)} \right) \\
Y(t) & = Y^*(t) \times \Ind{Y^*(\lfloor R \rfloor) = 0} \times \Ind{D(t) = 0} \ .
\end{aligned}\] The exogenous variables (errors)
\(U = \left( U_R, U_W, U_S, U_{H(t)}, U_{N(t)}, U_{A(t)}, U_{D(t)}, U_{Y^*(t)} \right)_{t = 1}^{T}\)
are mutually independent. Exposure status is a time-varying indicator;
let exposure history \(\bar A(t)\) be summarized as being ever-exposed
or not with \(\bar A(t) = \Ind{\sum_{k = 1}^t\Ind{A(k) = 1} > 0}\). The
outcome of interest is a survival outcome, so
\(Y^*(t - 1) = 1 \Rightarrow Y^*(t) = 1\). The observed outcome \(Y(t)\)
at time \(t\) is a function of true outcome status and the time points
delimiting left and right censoring. An abbreviated directed acyclic
graph (DAG) representing the causal relationships encoded in the
equations above is presented in Figure \ref{fig:dag}.

\begin{figure}
\caption{Directed acyclic graph representing the causal relationships encoded in the non-parametric structural equation model at time $t$.}
\label{fig:dag}
\begin{center}
\begin{tikzpicture}[>= stealth, auto, node distance = 2.25cm, semithick, fill=white, inner sep=0pt]
\tikzstyle{every state}=[shape = circle, align = center, draw = none]
\node[state] (H0) {$H(t - 1)$};
\node[state] (N0) [below right of=H0] {$N(t - 1)$};
\node[state] (A0) [below right of=N0] {$\bar A(t - 1)$};
\node[state] (D0) [below right of=A0] {$D(t - 1)$};
\node[state] (Y0) [below right of=D0] {$Y^*(t - 1)$};
\node[state] (W) [above right of=H0] {$W$};
\node[state] (H) [right of=H0, node distance = 4cm] {$H(t)$};
\node[state] (N) [below right of=H, node distance = 4cm] {$N(t)$};
\node[state] (A) [below right of=N, node distance = 4cm] {$\bar A(t)$};
\node[state] (D) [above right of=N, node distance = 3cm] {$D(t)$};
\node[state] (Y) [above right of=A, node distance = 3cm] {$Y^*(t)$};
\node[state] (S) [above right of=D, node distance = 3cm] {$S$};
%
\path[->] (H0) edge (H);
\path[->] (W) edge [bend right=10] (N);
\path[->] (N0) edge (N);
\path[->] (H) edge (N);
\path[->] (A0) edge (N);
\path[->] (W) edge [bend right=15] (A);
\path[->] (A0) edge (A);
\path[->] (N) edge (A);
\path[->] (W) edge (D);
\path[->] (A0) edge [bend left=15] (D);
\path[->] (D0) edge [bend right=15] (D);
\path[->] (Y0) edge [bend right=10] (D);
\path[->] (W) edge [bend left=5] (Y);
\path[->] (S) edge [bend left=30] (Y);
\path[->] (S) edge [bend left=15] (Y0);
\path[->] (H) edge (Y);
\path[->] (A) edge (Y);
\path[->] (D) edge (Y);
\path[->] (Y0) edge (Y);
\end{tikzpicture}
\end{center}
\end{figure}

@@

Parametric relationships between the variables were imposed in
simulations. For the \(n = 50\,000\) units over \(T = 20\) years, we
have:

@@

\begin{table}
\caption{Simulation parameters.}
\label{tab:params}
\begin{center}
\begin{tabular}{rrrrrr}
  \toprule
Parameter & Scenario 1 & Scenario 2 & Scenario 3 & Scenario 4 & Scenario 5 \\ 
  \midrule
$p_S$ & 0.10 & 0.10 & \textbf{0.20} & 0.10 & 0.10 \\ 
  $p_W$ & 0.20 & 0.20 & 0.20 & 0.20 & 0.20 \\ 
  $p_H$ & 0.30 & 0.30 & 0.30 & 0.30 & 0.30 \\ 
  $\beta_0^N$ & 3.00 & 3.00 & 3.00 & 3.00 & 3.00 \\ 
  $\beta_W^N$ & -0.10 & -0.10 & -0.10 & -0.10 & -0.10 \\ 
  $\beta_H^N$ & -0.50 & -0.50 & -0.50 & \textbf{-1.50} & -0.50 \\ 
  $\beta_A^N$ & -1.50 & -1.50 & -1.50 & -1.50 & -1.50 \\ 
  $\beta_0^A$ & -1.50 & -1.50 & -1.50 & -1.50 & -1.50 \\ 
  $\beta_W^A$ & -0.50 & -0.50 & -0.50 & -0.50 & -0.50 \\ 
  $\beta_A^A$ & 2.50 & 2.50 & 2.50 & 2.50 & 2.50 \\ 
  $\beta_0^D$ & -5.50 & -5.50 & -5.50 & -5.50 & -5.50 \\ 
  $\beta_W^D$ & 1.00 & 1.00 & 1.00 & 1.00 & 1.00 \\ 
  $\beta_{\bar A}^D$ & 0.50 & 0.50 & 0.50 & 0.50 & 0.50 \\ 
  $\beta_{\bar Y}^D$ & 0.50 & \textbf{2.00} & 0.50 & 0.50 & 0.50 \\ 
  $\beta_0^Y$ & -7.00 & -7.00 & -7.00 & -7.00 & \textbf{-6.00} \\ 
  $\beta_W^Y$ & 2.00 & 2.00 & 2.00 & 2.00 & 2.00 \\ 
  $\beta_A^Y$ & 0.25 & 0.25 & 0.25 & 0.25 & 0.25 \\ 
  $\beta_{\bar A}^Y$ & 0.20 & 0.20 & 0.20 & 0.20 & 0.20 \\ 
  $\beta_H^Y$ & 0.70 & 0.70 & 0.70 & \textbf{1.70} & 0.70 \\ 
  $\beta_S^Y$ & 0.30 & 0.30 & 0.30 & 0.30 & 0.30 \\ 
   \bottomrule
\end{tabular}
\end{center}
\end{table}

@@

The substantive question of interest is the causal effect of
occupational exposure to MWF on cancer incidence risk. Since
occupational MWF exposure occurs only when individuals are at work, we
define dynamic exposure regimes that depend on employment status. Under
rule \(a_0\), set \(D(t) = 0\), and set \(A(t) = 0\) while \(N(t) = 1\).
Under rule \(a_1\), set \(D(t) = 0\), and set \(A(t) = 1\) while
\(N(t) = 1\). Under both rules, we prevent censoring by death as if it
were intervenable. The causal effect is defined by contrasting the
survival function \(S_{a_1}(t) = 1 - \E{Y_{a_1}(t)}\) under rule \(a_1\)
to \(S_{a_0}(t) = 1 - \E{Y_{a_0}(t)}\) that under rule \(a_0\). Note
that this causal estimand is defined over \emph{a priori
counterfactuals} not observable in the real world (Frangakis and Rubin
2002). This approach is standard in epidemiologic studies, however.

The survival function expresses the probability that a person following
rule \(a\) is cancer-free at the end of time point \(t\). The expected
time until cancer under rule \(a\) is
\(\mu_{a} = \sum_0^K S_{a}(t) \, dt\). Our causal estimand is summary
measure \(\psi = \mu_{a_1} - \mu_{a_0}\), the difference in expected
time until event under two different interventions over 20 years of
follow-up under five different data generating scenarios. Bias will be
evaluated by comparing estimates of \(\psi\) to its true value in 250
simulations per scenario (the original analysis performed 500). The true
value will be calculated by simulating the full data for five hundred
thousand individuals (the original analysis used one million) with rules
\(a_0\) and \(a_1\) applied deterministically. Estimates of \(\psi\)
will be obtained by first estimating the survival curves \(S_a(t)\)
using two estimators: the inverse probability weighted Kaplan-Meier
estimator (WKM) and the Aalen-filtered WKM (AWKM). These survival
estimators are detailed in the following section.

@@

Figure \ref{fig:survival} presents the true survival curves as well as
the WKM and AWKM survival curves averaged over 250 replications for each
intervention rule and scenario. Qualitatively, the WKM estimator
consistently over-estimates survival whereas the the AWKM survival curve
is much closer to the truth. The bias of the AWKM survival estimate
appears to be larger earlier in follow-up and smaller as follow-up
extends forward. The bias of both the the WKM and the AWKM survival
curves appears largest in Scenario 5.

@@

Table \ref{tab:survival} presents true and estimated average cancer-free
survival times under each intervention rule and scenario. Table
\ref{tab:differences} presents differences in survival time contrasting
rule \(a_1\) to rule \(a_0\). Table \ref{tab:bias} presents estimates of
the bias of the WKM and AWKM estimators for \(\psi\), the difference in
average cancer-free survival time over 20 years of follow-up. These
numeric results are consistent with the qualitative interpretations of
Figure \ref{fig:survival}. The WKM estimator over-estimates the
difference in cancer-free survival time, resulting in bias toward the
null, whereas the AWKM estimator under-estimates the cancer-free
survival, resulting in bias away from the null. In every scenario, the
bias of the WKM-derived contrast is several times larger in magnitude
than that of the AWKM-derived contrast. The bias of both estimators is
greatest for Scenario 5.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Estimating survival in left filtered data},
  pdfauthor={Kevin Chen},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=2.54cm]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\PassOptionsToPackage{no-math}{fontspec}

@@

\title{Estimating survival in left filtered data}
\author{Kevin Chen}
\date{Stat 256: Causal Inference (Fall 2021)}

@@

A central concern in the analysis of occupational cohorts is the
potential for the healthy worker survivor effect (HWSE), the phenomenon
by which healthy individuals remain at work, while less healthy
individuals leave work -- possibly in response to exposure-related
health decline. In the presence of the HWSE, those with the highest
cumulative occupational exposures are also those who are less at risk of
disease. Thus, standard measures of association would show an inverse
relationship between occupational exposure and poor health outcomes
(Arrighi and Hertz-Picciotto 1994). The HWSE is an example of
time-varying confounding affected by past exposure. Previous studies
have attempted to assess the presence of the HWSE in observed data by
assessing so-called path-specific associations using Cox proportional
hazards modeling (Naimi et al. 2013; Garcia et al. 2017). However, these
measures of associations were themselves subject to the confounding
structures they sought to characterize.

If conditional sequential ignorability of exposure and censoring status
at each point in follow-up and positivity can be attained given
covariate history, then causal methods can be applied to account for the
HWSE. Past studies have applied causal methods capable of accounting for
time-varying confounding affected by past exposure to the study of MWF
exposures and cancer mortality outcomes in the UAW-GM Cohort Study
(Garcia et al. 2018; Izano et al. 2019), but the study of cancer
incidence outcomes is further problematized by the incomplete
observation of cancer incidence outcomes at every point of follow-up
over the study period. Nonetheless, we wish to make inferences about the
carcinogenicity of MWF exposure over an individual's lifetime starting
upon entry into the workforce. The UAW-GM Cohort included those hired
roughly between 1938 and 1985. However, cancer incidence reporting at
the Michigan Cancer Registry did not begin until 1985. Hence, our
observed cohort data is subject to \emph{left filtering}: cancer
incidence is the outcome of interest, but before 1985, both cancer
incidence status and time of cancer incidence are unknown. Observation
of the complete cancer incidence outcome vector over the study period
was conditional on an individual surviving to 1985 cancer-free.

In the presence of the HWSE, left filtering implies outcome
misclassification that is informative of true cancer status. As part of
her dissertation research, Izano (2017) conducted a quantitative bias
analysis for the estimation of survival curves using left filtered data
in the presence of HWSE. She simulated data compatible with the HWSE and
estimated cancer-free survival curves using an inverse probability of
treatment and censoring weighted Kaplan-Meier (WKM) estimator and the
WKM with an Aalen filter for left-filtering (AWKM) (Andersen et al.
1993; Xie and Liu 2005). Data were simulated under five different
scenarios. The survival curves under the following interventions were
computed or estimated in 250 replicates: (1) always exposed at work with
no censoring due to death and (2) never exposed at work with no
censoring due to death. For each intervention, three survival curves
were produced: (1) the true survival curve, (2) the WKM survival curve,
and (3) AWKM survival curve. Estimator bias was evaluated by comparing
the average of 250 estimates of to the average difference in cancer-free
survival under the two rules to the true average survival difference
when exposure and censoring were controlled deterministically.

The present project replicates the simulation and bias analyses
presented in Chapter 3 of Izano (2017), embeds the problem in the
non-parametric structural causal approach of Pearl (1995), and applies
the WKM and AWKM estimators to the estimation of colon cancer survival
under exposure/censoring rules in the UAW-GM Cohort.

@@

The UAW-GM Cohort data included person-year level exposure, outcome, and
covariate data starting at hire. To emulate the shape of the data for
this longitudinal cohort, we considered 20 years of data over time
indexed by years since hire. Notation representing the variables of
interest are presented in Table \ref{tab:variables}. The causal model
represents hypothetical relationships between variables over time
compatible with the theory underlying the HWSE in longitudinal
occupational cohort studies. At each time point, the effect of
cumulative exposure \(\bar A(t)\) on cancer incidence \(Y^(t)\) is
confounded by a path through employment status \(N(t)\) and underlying
health \(H(t)\) as well as the path through past exposure
\(\bar A(t - 1)\) and vital status \(D(t)\). These paths follow
straightforward logic: occupational exposure depends upon employment
status and past exposure; mortality status is affected by past exposure
and cancer history. Confounding by baseline covariates \(W\) is assumed
throughout.

@@

Assume we have \(n = 50\,000\) iid units in \(X\) with
\[X_i(t) = \left(R_i = 0, W_i, S_i, \bar H_i(t), \bar N_i(t), \bar A_i(t), \bar a_i(t), \bar Y^*_i(t) = \bar Y_i(t) \right).\]
The bar notation to indicates variable history where
\(\bar X_i(t) = \left(X_i(k)\right)_{k = 1}^t\). In the case of
exposure, \(\bar A_i(t)\) is the indicator of whether unit \(i\) was
ever exposed through time \(t\). Note that true cancer status \(Y^*(t)\)
not observed until \(t \ge R\), after the start of the registry. Call
\(X\) the full data, where we have \(R = 0\) for all. In the observed
data \(X^\text{obs}\), we cannot assume \(R = 0\) for all;
susceptibility \(S\) and underlying health status \(H\) are not known:
\[X^\text{obs}_i(t) = \left(R_i, W_i, \bar N_i(t), \bar A_i(t), \bar a_i(t), \bar Y_i(t) \right).\]
Under the causal model, we assume the following non-parametric
structural equations: \[\begin{aligned}
R     & = f_R \left( U_R \right) \\
W     & = f_W \left( U_W \right) \\
S     & = f_S \left( U_S \right) \\
H(t)  & = f_{H(t)} \left( H(t - 1), U_{H(t)} \right) \\
N(t)  & = f_{N(t)} \left( W, N(t - 1), H(t), A(t - 1), U_{N(t)} \right) \\
A(t)  & = f_{A(t)} \left( W, \bar A(t - 1), N(t), U_{A(t)} \right) \\
D(t)  & = f_{D(t)} \left( W, \bar A(t - 1), D(t - 1), Y^*(t - 1), U_{D(t)} \right) \\
Y^*(t)& = f_{Y^*(t)} \left(W, S, H(t), \bar A (t), D(t), Y^*(t - 1), U_{Y^*(t)} \right) \\
Y(t) & = Y^*(t) \times \Ind{Y^*(\lfloor R \rfloor) = 0} \times \Ind{D(t) = 0} \ .
\end{aligned}\] The exogenous variables (errors)
\(U = \left( U_R, U_W, U_S, U_{H(t)}, U_{N(t)}, U_{A(t)}, U_{D(t)}, U_{Y^*(t)} \right)_{t = 1}^{T}\)
are mutually independent. Exposure status is a time-varying indicator;
let exposure history \(\bar A(t)\) be summarized as being ever-exposed
or not with \(\bar A(t) = \Ind{\sum_{k = 1}^t\Ind{A(k) = 1} > 0}\). The
outcome of interest is a survival outcome, so
\(Y^*(t - 1) = 1 \Rightarrow Y^*(t) = 1\). The observed outcome \(Y(t)\)
at time \(t\) is a function of true outcome status and the time points
delimiting left and right censoring. An abbreviated directed acyclic
graph (DAG) representing the causal relationships encoded in the
equations above is presented in Figure \ref{fig:dag}.

\begin{figure}
\caption{Directed acyclic graph representing the causal relationships encoded in the non-parametric structural equation model at time $t$.}
\label{fig:dag}
\begin{center}
\begin{tikzpicture}[>= stealth, auto, node distance = 2.25cm, semithick, fill=white, inner sep=0pt]
\tikzstyle{every state}=[shape = circle, align = center, draw = none]
\node[state] (H0) {$H(t - 1)$};
\node[state] (N0) [below right of=H0] {$N(t - 1)$};
\node[state] (A0) [below right of=N0] {$\bar A(t - 1)$};
\node[state] (D0) [below right of=A0] {$D(t - 1)$};
\node[state] (Y0) [below right of=D0] {$Y^*(t - 1)$};
\node[state] (W) [above right of=H0] {$W$};
\node[state] (H) [right of=H0, node distance = 4cm] {$H(t)$};
\node[state] (N) [below right of=H, node distance = 4cm] {$N(t)$};
\node[state] (A) [below right of=N, node distance = 4cm] {$\bar A(t)$};
\node[state] (D) [above right of=N, node distance = 3cm] {$D(t)$};
\node[state] (Y) [above right of=A, node distance = 3cm] {$Y^*(t)$};
\node[state] (S) [above right of=D, node distance = 3cm] {$S$};
%
\path[->] (H0) edge (N0);
\path[->] (N0) edge (A0);
\path[->] (A0) edge [bend right=20] (Y0);
\path[->] (H0) edge (H);
\path[->] (W) edge [bend right=10] (N);
\path[->] (N0) edge (N);
\path[->] (H) edge (N);
\path[->] (A0) edge (N);
\path[->] (W) edge [bend right=15] (A);
\path[->] (A0) edge (A);
\path[->] (N) edge (A);
\path[->] (W) edge (D);
\path[->] (A0) edge [bend left=15] (D);
\path[->] (D0) edge [bend right=15] (D);
\path[->] (Y0) edge [bend right=10] (D);
\path[->] (W) edge [bend left=5] (Y);
\path[->] (S) edge [bend left=30] (Y);
\path[->] (S) edge [bend left=15] (Y0);
\path[->] (H) edge (Y);
\path[->] (A) edge (Y);
\path[->] (D) edge (Y);
\path[->] (Y0) edge (Y);
\end{tikzpicture}
\end{center}
\end{figure}

Figure \ref{fig:dag} clarifies our conceptualization of the HWSE. At
each time point \(t\), the cumulative effect of exposure \(\bar A(t)\)
on cancer incidence \(Y^*(t)\) is confounded by employment status
\(N(t)\) through the backdoor path
\(\bar A(t) \leftarrow N(t) \leftarrow H(t) \rightarrow Y^*(t)\). In the
absence of observed data on health status \(H(t)\), an analyst may be
tempted to condition on employment status \(N(t)\), but doing so would
introduce collider bias while blocking the causal path between past
exposure \(\bar A(t - 1)\) and cancer \(Y^*(t)\). Furthermore, an
analysis starting at an arbitrary time point after the time origin
(hire) would be tantamount to conditioning on those still alive at that
time, which would also result in both collider bias and the conditioning
on nodes on the causal path between the exposure and outcome of
interest.

@@

To generate data compatible with our structural causal model, we imposed
parametric relationships between the variables. For the \(n = 50\,000\)
units over \(T = 20\) years, we have:

@@

\begin{table}[h]
\caption{Simulation parameters.}
\label{tab:params}
\begin{center}
\begin{tabular}{rrrrrr}
  \toprule
Parameter & Scenario 1 & Scenario 2 & Scenario 3 & Scenario 4 & Scenario 5 \\ 
  \midrule
$p_S$ & 0.10 & 0.10 & \textbf{0.20} & 0.10 & 0.10 \\ 
  $p_W$ & 0.20 & 0.20 & 0.20 & 0.20 & 0.20 \\ 
  $p_H$ & 0.30 & 0.30 & 0.30 & 0.30 & 0.30 \\ 
  $\beta_0^N$ & 3.00 & 3.00 & 3.00 & 3.00 & 3.00 \\ 
  $\beta_W^N$ & -0.10 & -0.10 & -0.10 & -0.10 & -0.10 \\ 
  $\beta_H^N$ & -0.50 & -0.50 & -0.50 & \textbf{-1.50} & -0.50 \\ 
  $\beta_A^N$ & -1.50 & -1.50 & -1.50 & -1.50 & -1.50 \\ 
  $\beta_0^A$ & -1.50 & -1.50 & -1.50 & -1.50 & -1.50 \\ 
  $\beta_W^A$ & -0.50 & -0.50 & -0.50 & -0.50 & -0.50 \\ 
  $\beta_A^A$ & 2.50 & 2.50 & 2.50 & 2.50 & 2.50 \\ 
  $\beta_0^D$ & -5.50 & -5.50 & -5.50 & -5.50 & -5.50 \\ 
  $\beta_W^D$ & 1.00 & 1.00 & 1.00 & 1.00 & 1.00 \\ 
  $\beta_{\bar A}^D$ & 0.50 & 0.50 & 0.50 & 0.50 & 0.50 \\ 
  $\beta_{\bar Y}^D$ & 0.50 & \textbf{2.00} & 0.50 & 0.50 & 0.50 \\ 
  $\beta_0^Y$ & -7.00 & -7.00 & -7.00 & -7.00 & \textbf{-6.00} \\ 
  $\beta_W^Y$ & 2.00 & 2.00 & 2.00 & 2.00 & 2.00 \\ 
  $\beta_A^Y$ & 0.25 & 0.25 & 0.25 & 0.25 & 0.25 \\ 
  $\beta_{\bar A}^Y$ & 0.20 & 0.20 & 0.20 & 0.20 & 0.20 \\ 
  $\beta_H^Y$ & 0.70 & 0.70 & 0.70 & \textbf{1.70} & 0.70 \\ 
  $\beta_S^Y$ & 0.30 & 0.30 & 0.30 & 0.30 & 0.30 \\ 
   \bottomrule
\end{tabular}
\end{center}
\end{table}

@@

The substantive question of interest was the causal effect of
occupational exposure to MWF on cancer incidence risk. Since
occupational MWF exposure occurs only when individuals are at work, we
defined dynamic exposure regimes that depend on employment status. Under
rule \(a_0\), set \(D(t) = 0\), and set \(A(t) = 0\) while \(N(t) = 1\).
Under rule \(a_1\), set \(D(t) = 0\), and set \(A(t) = 1\) while
\(N(t) = 1\). Under both rules, we prevented censoring by death as if it
were intervenable. The causal effect was defined by contrasting the
survival function \(S_{a_1}(t) = 1 - \E{Y_{a_1}(t)}\) under rule \(a_1\)
to \(S_{a_0}(t) = 1 - \E{Y_{a_0}(t)}\) that under rule \(a_0\). Note
that this causal estimand was defined over \emph{a priori
counterfactuals} not observable in the real world (Frangakis and Rubin
2002). This approach is standard in epidemiologic studies, however.

The survival function expresses the probability that a person following
rule \(a\) is cancer-free at the end of time point \(t\). The expected
time until cancer under rule \(a\) is
\(\mu_{a} = \sum_0^K S_{a}(t) \, dt\). The parameter used in the
estimation of bias was the summary measure
\(\psi = \mu_{a_1} - \mu_{a_0}\), the difference in expected time until
event under two different interventions over 20 years of follow-up under
five different data generating scenarios. Bias was evaluated by
comparing estimates of \(\psi\) to its true value in 250 simulations per
scenario (the original analysis performed 500). The true value was
calculated by simulating the full data for five hundred thousand
individuals (the original analysis used one million) with rules \(a_0\)
and \(a_1\) applied deterministically. Estimates of \(\psi\) were
obtained by first estimating the survival curves \(S_a(t)\) using two
estimators: the inverse probability weighted Kaplan-Meier estimator
(WKM) and the Aalen-filtered WKM (AWKM). These survival estimators are
detailed in the following section.

@@

Figure \ref{fig:survival} presents the true survival curves as well as
the WKM and AWKM survival curves averaged over 250 replications for each
intervention rule and scenario. Qualitatively, the WKM estimator
consistently over-estimated survival whereas the the AWKM survival curve
was much closer to the truth. The bias of the AWKM survival estimator
appeared to be larger in earlier follow-up and smaller for later
follow-up time points. The bias of both the the WKM and the AWKM
estimators appeared largest in Scenario 5.

@@

Table \ref{tab:survival} presents true and estimated average cancer-free
survival times under each intervention rule and scenario. Table
\ref{tab:differences} presents differences in survival time contrasting
rule \(a_1\) to rule \(a_0\). Table \ref{tab:bias} presents estimates of
the bias of the WKM and AWKM estimators for \(\psi\), the difference in
average cancer-free survival time over 20 years of follow-up. These
numeric results were consistent with the qualitative interpretations of
Figure \ref{fig:survival}. The WKM estimator over-estimated the
difference in cancer-free survival time, resulting in bias toward the
null, whereas the AWKM estimator under-estimates the cancer-free
survival, resulting in bias away from the null. In every scenario, the
bias of the WKM-derived contrast was several times larger in magnitude
than that of the AWKM-derived contrast. The bias of both estimators was
greatest for Scenario 5.
